---
title: "PA1_template.Rmd"
output: html_document
---

## Loading and preprocessing the data

```{r loading}
data <- read.csv("activity.csv")
data$date <- as.Date(paste(data$date))
```

## Histogram of Average steps taken each day before imputing

The following code chunk creates a histogram of the total number of steps taken per day
and calculates mean and median of that data.

```{r histogram_1}
x <- tapply(data$steps, data$date, sum)
hist(x, main=paste("Average steps taken each day before Imputing missing data"), xlab="Steps",ylab="#Days")
y <- as.numeric(paste(x))
isNA <- is.na(y)
y <- y[!isNA]
mean(y)
median(y)
```

As you can see above:

1. Mean: 10766.19
2. Median: 10765

## 5 minute interval time series plot

```{r time series plot}
compcases <- complete.cases(data)
datacomplete <- data[compcases,]

timeseries <- tapply(datacomplete$steps, datacomplete$interval, sum)
plot(timeseries, type = "l", xlab="5min interval", ylab="steps", main="Time Series plot of steps versus time interval")
order <- order(timeseries)
timeseries[order[length(order)]]
```

As you can see, there is a massive peak relatively early in the day.
The five minute interval with the maximum number of steps is interval 835, with 10927 steps.

## Imputing missing Values

I choose to Impute missing step data by averaging over the corresponding 5 minute interval step data of all
the other days.

```{r imputing}
isNA <- is.na(data$steps)
sum(isNA)

totals <- tapply(datacomplete$steps, datacomplete$interval, sum)
days <- unique(data$date)
days <- length(days)
averages <- totals/days

data2 <- data
data2$averages <- averages
data2$steps[!compcases] <- data2$averages

head(data)
head(data2)
```

Here, head(data) and head(data2) have been printed to showcase how averages populate previously "NA"'d values.

## Histogram of Average steps taken each day after imputing

```{r histogram_2}
x <- tapply(data2$steps, data2$date, sum)
hist(x, main=paste("Average steps taken each day after Imputing missing data"), xlab="Steps",ylab="#Days")
y <- as.numeric(paste(x))
isNA <- is.na(y)
y <- y[!isNA]
mean(y)
median(y)
```

1. Mean: 10581.01
2. Median: 10395

There are now a couple more days in the 5000-10000 steps range.

## Weekend/Weekday split

```{r morestuff}
data2$weekdays <- weekdays(data2$date)
data2$weekend <- FALSE
data2$weekend <- data2$weekday %in% c("Saturday","Sunday")
head(data2)
```

```{r moregraphs}
par(mfrow=c(2,1))
weekdays <- subset(data2, data2$weekend==FALSE)
weekends <- subset(data2, data2$weekend==TRUE)
weekday_averages <- tapply(weekdays$steps, weekdays$interval, sum)
weekday_averages <- weekday_averages/length(unique(weekdays$date))

weekend_averages <- tapply(weekends$steps, weekends$interval, sum)
weekend_averages <- weekend_averages/length(unique(weekends$date))

plot(weekend_averages, type="l", xlab="Interval", ylab="Number of steps", main="weekend")
plot(weekday_averages, type="l", xlab="Interval", ylab="Number of steps", main="weekday")
```

The differences are quite evident. People are mostly not walking during the day on weekdays. On weekends, they spend the entire day walking.